---
title: "STA 9750 Mini-Project #01"
subtitle: "Assessing the Impact of SFFA on Campus Diversity One-Year Later"
author: "Kiran Narsipur"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(glue)
library(gt)
library(skimr)
library(scales)
library(stringr)
```

# Introduction

This report analyzes IPEDS enrollment data to explore how campus diversity changed before and after the Supreme Court's 2023 decision in *Students for Fair Admissions v. Harvard* (SFFA). The goal is not to make strong causal claims, but rather to use descriptive analysis to see whether 2024 enrollment patterns look materially different from prior years and whether those changes are unusually large relative to benchmark institutions.

I begin by acquiring and preparing the IPEDS data, then create two institutional indicators required for the assignment: one for CUNY institutions and another for California public institutions. After that, I perform an initial exploratory data analysis and answer a set of descriptive questions using inline values and formatted tables. I conclude with an Op-Ed written from the perspective of a college president.

# Data Acquisition

## Task 1: Acquire the IPEDS data

```{r}
#| label: acquire-data
#' Acquire IPEDS Data for MP#01
#'
#' This function will acquire and standardize all data for MP#01
#' from IPEDS (https://nces.ed.gov/ipeds/datacenter/DataFiles.aspx)
#'
#' We're starting in 2010 as the data seems to be reasonably complete
#' after that point.
acquire_ipeds_data <- function(start_year = 2010, end_year = 2024) {
  data_dir <- file.path("data", "mp01")

  if (!dir.exists(data_dir)) {
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
  }

  YEARS <- seq(start_year, end_year)

  EFA_ALL <- map(YEARS, function(yy) {
    if (yy <= 2022) {
      ef_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
    } else {
      ef_url <- glue(
        "https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv"
      )
    }

    ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))

    if (!file.exists(ef_file)) {
      message(glue("Downloading Enrollment Data for {yy} from {ef_url}"))
      download.file(ef_url, destfile = ef_file, quiet = TRUE, mode = "wb")
    }

    read_csv(ef_file, show_col_types = FALSE) |>
      mutate(
        year = yy,
        enrollment_m_aian = EFAIANM,
        enrollment_f_aian = EFAIANW,
        enrollment_m_asia = EFASIAM,
        enrollment_f_asia = EFASIAW,
        enrollment_m_bkaa = EFBKAAM,
        enrollment_f_bkaa = EFBKAAW,
        enrollment_m_hisp = EFHISPM,
        enrollment_f_hisp = EFHISPW,
        enrollment_m_nhpi = EFNHPIM,
        enrollment_f_nhpi = EFNHPIW,
        enrollment_m_whit = EFWHITM,
        enrollment_f_whit = EFWHITW,
        enrollment_m_2mor = EF2MORM,
        enrollment_f_2mor = EF2MORW,
        enrollment_m_unkn = EFUNKNM,
        enrollment_f_unkn = EFUNKNW,
        enrollment_m_nral = EFNRALM,
        enrollment_f_nral = EFNRALW
      ) |>
      filter(
        (EFALEVEL %in% c(2, 12)) | (LINE %in% c(1, 15))
      ) |>
      mutate(
        level = case_when(
          EFALEVEL == 2 ~ "all undergraduate",
          EFALEVEL == 12 ~ "all graduate",
          LINE %in% c(1, 15) ~ "first year undergrad"
        )
      ) |>
      select(
        institution_id = UNITID,
        year,
        level,
        starts_with("enrollment_")
      ) |>
      group_by(institution_id, year, level) |>
      summarize(across(starts_with("enrollment_"), sum), .groups = "drop")
  }) |>
    bind_rows()

  DESC_ALL <- map(YEARS, function(yy) {
    if (yy <= 2022) {
      hd_url <- glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
    } else {
      hd_url <- glue(
        "https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv"
      )
    }

    hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))

    if (!file.exists(hd_file)) {
      message(glue("Downloading Institutional Descriptions for {yy} from {hd_url}"))
      download.file(hd_url, destfile = hd_file, quiet = TRUE, mode = "wb")
    }

    suppressWarnings(
      read_csv(
        hd_file,
        show_col_types = FALSE,
        locale = locale(encoding = if_else(yy == 2024, "utf-8", "windows-1252"))
      ) |>
        mutate(year = yy, INSTNM = INSTNM) |>
        select(
          institution_id = UNITID,
          institution_name = INSTNM,
          state = STABBR,
          year
        )
    )
  }) |>
    bind_rows()

  inner_join(
    EFA_ALL,
    DESC_ALL,
    join_by(institution_id == institution_id, year == year)
  )
}

IPEDS <- acquire_ipeds_data()
```

The code above downloads and prepares the IPEDS enrollment and institution description files and stores them in a local `data/mp01` folder so that future renders do not repeatedly hit the IPEDS servers.

# Data Cleaning and Preparation

## Task 2: Identify CUNY schools

```{r}
#| label: create-indicators
IPEDS <- IPEDS |>
  mutate(
    is_cuny = str_detect(institution_name, "CUNY"),
    is_calpublic =
      state == "CA" & (
        str_detect(institution_name, "University of California") |
          str_detect(institution_name, "California State University") |
          str_detect(institution_name, "California Polytechnic State University") |
          str_detect(institution_name, "California State Polytechnic University")
      )
  )
```

## Task 3: Identify UC and Cal State schools

The following table shows the California public institutions identified by the logic above.

```{r}
#| label: calpublic-list
IPEDS |>
  filter(is_calpublic) |>
  distinct(institution_name) |>
  arrange(institution_name) |>
  gt() |>
  tab_header(
    title = "California Public Institutions Identified in the IPEDS Data",
    subtitle = "Institutions flagged as University of California or California State system schools"
  ) |>
  cols_label(institution_name = "Institution Name") |>
  tab_source_note("Source: IPEDS institutional descriptions, filtered using string-matching rules in this report.")
```

# Additional Variables for Analysis

To make later tasks cleaner, I create a few summary variables that combine the male and female counts into total counts by race/ethnicity, as well as an overall enrollment total.

```{r}
#| label: derived-variables
IPEDS <- IPEDS |>
  mutate(
    total_aian = enrollment_m_aian + enrollment_f_aian,
    total_asia = enrollment_m_asia + enrollment_f_asia,
    total_bkaa = enrollment_m_bkaa + enrollment_f_bkaa,
    total_hisp = enrollment_m_hisp + enrollment_f_hisp,
    total_nhpi = enrollment_m_nhpi + enrollment_f_nhpi,
    total_whit = enrollment_m_whit + enrollment_f_whit,
    total_2mor = enrollment_m_2mor + enrollment_f_2mor,
    total_unkn = enrollment_m_unkn + enrollment_f_unkn,
    total_nral = enrollment_m_nral + enrollment_f_nral,
    total_female = enrollment_f_aian + enrollment_f_asia + enrollment_f_bkaa +
      enrollment_f_hisp + enrollment_f_nhpi + enrollment_f_whit +
      enrollment_f_2mor + enrollment_f_unkn + enrollment_f_nral,
    total_male = enrollment_m_aian + enrollment_m_asia + enrollment_m_bkaa +
      enrollment_m_hisp + enrollment_m_nhpi + enrollment_m_whit +
      enrollment_m_2mor + enrollment_m_unkn + enrollment_m_nral,
    total_enrollment = total_female + total_male,
    total_nonwhite_nonasian = total_aian + total_bkaa + total_hisp + total_nhpi + total_2mor,
    share_white = total_whit / total_enrollment,
    share_female = total_female / total_enrollment,
    share_nhpi = total_nhpi / total_enrollment,
    share_nonwhite_nonasian = total_nonwhite_nonasian / total_enrollment
  )
```

# Initial EDA

## Task 4: Initial EDA pass

I first check the structure and types of the data and then review univariate summaries.

```{r}
#| label: glimpse-ipeds
glimpse(IPEDS)
```

The `glimpse()` output shows that the data has one row per institution-year-level combination, with numeric enrollment counts and character fields for institution name, state, and level. The newly created CUNY and California-public indicators are logical variables, which is exactly what I want for later filtering tasks.

```{r}
#| label: skim-ipeds
skim(IPEDS)
```

The `skim()` summary confirms that the enrollment columns are numeric and that the key institution identifiers are stored consistently. The distribution summaries also show that enrollment values vary widely across institutions, which is expected because this dataset combines very small colleges with very large public university systems. That variation is an important reason to be careful when ranking institutions by proportional changes, since small schools can show large swings from a small baseline.

# Exploratory Questions: Inline Values

```{r}
#| label: inline-objects
# Distinct institutions
n_institutions <- n_distinct(IPEDS$institution_id)

# Baruch helpers
baruch_name <- IPEDS |>
  filter(str_detect(institution_name, "Baruch")) |>
  distinct(institution_name) |>
  slice(1) |>
  pull(institution_name)

baruch_2024_grad <- IPEDS |>
  filter(year == 2024, str_detect(institution_name, "Baruch"), level == "all graduate") |>
  summarize(val = sum(total_enrollment, na.rm = TRUE)) |>
  pull(val)

baruch_2024_total <- IPEDS |>
  filter(year == 2024, str_detect(institution_name, "Baruch"), level %in% c("all undergraduate", "all graduate")) |>
  summarize(val = sum(total_enrollment, na.rm = TRUE)) |>
  pull(val)

female_2019_top <- IPEDS |>
  filter(year == 2019, level %in% c("all undergraduate", "all graduate")) |>
  group_by(institution_name) |>
  summarize(total_female_students = sum(total_female, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total_female_students)) |>
  slice(1)

female_2019_top_name <- female_2019_top$institution_name[[1]]
female_2019_top_n <- female_2019_top$total_female_students[[1]]

nhpi_2024_top <- IPEDS |>
  filter(year == 2024, level == "first year undergrad", total_enrollment > 1000) |>
  mutate(nhpi_share = total_nhpi / total_enrollment) |>
  arrange(desc(nhpi_share)) |>
  slice(1)

nhpi_2024_top_name <- nhpi_2024_top$institution_name[[1]]
nhpi_2024_top_share <- nhpi_2024_top$nhpi_share[[1]]
nhpi_2024_top_total <- nhpi_2024_top$total_enrollment[[1]]
```

## Task 5: Inline answers

There are `r comma(n_institutions)` distinct institutions represented in this dataset.

In 2024, `r baruch_name` enrolled `r comma(baruch_2024_grad)` graduate students.

In 2024, `r baruch_name` enrolled `r comma(baruch_2024_total)` total students when undergraduate and graduate students are counted without double-counting the first-year subgroup.

In 2019, the institution with the highest total number of enrolled female students was `r female_2019_top_name`, with `r comma(female_2019_top_n)` female students.

Among institutions with more than 1,000 first-year undergraduate students in 2024, `r nhpi_2024_top_name` had the highest proportion of Native Hawaiian or Pacific Islander first-year undergraduates at `r percent(nhpi_2024_top_share, accuracy = 0.01)`. That institution enrolled `r comma(nhpi_2024_top_total)` first-year undergraduates overall.

# Exploratory Questions: Table Answers

## Task 6A: Which 5 states had the highest number of graduate students?

```{r}
#| label: tbl-top-grad-states
top_grad_states <- IPEDS |>
  filter(year == 2024, level == "all graduate") |>
  group_by(state) |>
  summarize(graduate_students = sum(total_enrollment, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(graduate_students)) |>
  slice_head(n = 5)

top_grad_states |>
  gt() |>
  tab_header(
    title = "Top 5 States by Graduate Enrollment",
    subtitle = "Graduate students across all institutions in 2024"
  ) |>
  cols_label(
    state = "State",
    graduate_students = "Graduate Students"
  ) |>
  fmt_number(columns = graduate_students, decimals = 0) |>
  tab_source_note("Source: IPEDS 2024 graduate enrollment records.")
```

The largest graduate student populations are concentrated in large states with extensive public and private higher-education systems. This is not surprising, but it provides a useful benchmark for understanding the scale of graduate education nationally.

## Task 6B: In 2024, how many first-year undergraduates were enrolled at CUNY colleges, and which colleges did they attend?

```{r}
#| label: tbl-cuny-fyug
cuny_fy_2024 <- IPEDS |>
  filter(year == 2024, level == "first year undergrad", is_cuny) |>
  group_by(institution_name) |>
  summarize(first_year_students = sum(total_enrollment, na.rm = TRUE), .groups = "drop") |>
  mutate(share_of_cuny = first_year_students / sum(first_year_students)) |>
  arrange(desc(first_year_students))

cuny_total_2024 <- sum(cuny_fy_2024$first_year_students)

cuny_fy_2024 |>
  gt() |>
  tab_header(
    title = "First-Year Undergraduate Enrollment at CUNY Colleges",
    subtitle = glue("CUNY enrolled {comma(cuny_total_2024)} first-year undergraduates in 2024")
  ) |>
  cols_label(
    institution_name = "CUNY Institution",
    first_year_students = "First-Year Students",
    share_of_cuny = "% of CUNY Total"
  ) |>
  fmt_number(columns = first_year_students, decimals = 0) |>
  fmt_percent(columns = share_of_cuny, decimals = 1) |>
  tab_source_note("Source: IPEDS 2024 first-year undergraduate enrollment records.")
```

This table shows how the entering CUNY class is distributed across colleges rather than being evenly spread across the system. Larger campuses account for a much bigger share of first-year students than smaller specialized campuses.

## Task 6C: How has Baruch’s total undergraduate enrollment changed over the study period?

```{r}
#| label: tbl-baruch-ug-trend
baruch_ug_trend <- IPEDS |>
  filter(str_detect(institution_name, "Baruch"), level == "all undergraduate") |>
  arrange(year) |>
  mutate(
    yoy_change = total_enrollment - lag(total_enrollment),
    yoy_pct_change = total_enrollment / lag(total_enrollment) - 1
  ) |>
  select(year, total_enrollment, yoy_change, yoy_pct_change)

baruch_ug_trend |>
  gt() |>
  tab_header(
    title = "Baruch Undergraduate Enrollment Over Time",
    subtitle = "Total undergraduate enrollment and year-over-year change"
  ) |>
  cols_label(
    year = "Year",
    total_enrollment = "Undergraduate Enrollment",
    yoy_change = "YoY Change",
    yoy_pct_change = "YoY % Change"
  ) |>
  fmt_number(columns = c(total_enrollment, yoy_change), decimals = 0) |>
  fmt_percent(columns = yoy_pct_change, decimals = 1) |>
  tab_source_note("Source: IPEDS undergraduate enrollment records for Baruch College.")
```

This table helps separate true enrollment shifts from normal year-to-year variation. Looking at multiple years also makes it easier to judge whether 2024 appears unusual or simply part of a broader trend already underway.

## Task 6D: At what 5 institutions did the fraction of white students decrease the most from 2010 to 2020?

```{r}
#| label: tbl-white-share-decline
white_share_change <- IPEDS |>
  filter(level == "all undergraduate", year %in% c(2010, 2020), total_enrollment >= 5000) |>
  select(institution_id, institution_name, year, total_enrollment, share_white) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_enrollment, share_white),
    names_sep = "_"
  ) |>
  filter(!is.na(share_white_2010), !is.na(share_white_2020)) |>
  mutate(change_white_share = share_white_2020 - share_white_2010) |>
  arrange(change_white_share) |>
  slice_head(n = 5) |>
  select(institution_name, share_white_2010, share_white_2020, change_white_share)

white_share_change |>
  gt() |>
  tab_header(
    title = "Institutions with the Largest Declines in White Undergraduate Share",
    subtitle = "Restricted to institutions with at least 5,000 undergraduates in both comparison years"
  ) |>
  cols_label(
    institution_name = "Institution",
    share_white_2010 = "White Share (2010)",
    share_white_2020 = "White Share (2020)",
    change_white_share = "Change"
  ) |>
  fmt_percent(columns = c(share_white_2010, share_white_2020, change_white_share), decimals = 1) |>
  tab_source_note("Source: IPEDS undergraduate enrollment data for 2010 and 2020.")
```

I filtered out smaller institutions before ranking changes in white share because small campuses can show dramatic proportional swings that do not reflect broader structural change. This makes the comparison more meaningful.

## Task 6E: In which 3 states did the fraction of female undergraduates increase the most from 2010 to 2024?

```{r}
#| label: tbl-female-share-states

valid_states <- c(state.abb, "DC")

female_share_states <- IPEDS |>
  filter(
    level == "all undergraduate",
    year %in% c(2010, 2024),
    state %in% valid_states
  ) |>
  group_by(state, year) |>
  summarize(
    total_female = sum(total_female, na.rm = TRUE),
    total_students = sum(total_enrollment, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(female_share = total_female / total_students) |>
  select(state, year, female_share) |>
  pivot_wider(
    names_from = year,
    values_from = female_share,
    names_prefix = "share_"
  ) |>
  mutate(change = share_2024 - share_2010) |>
  arrange(desc(change)) |>
  slice_head(n = 3)

female_share_states |>
  gt() |>
  tab_header(
    title = "States with the Largest Increase in Female Undergraduate Share",
    subtitle = "Change from 2010 to 2024"
  ) |>
  cols_label(
    state = "State",
    share_2010 = "Female Share (2010)",
    share_2024 = "Female Share (2024)",
    change = "Change"
  ) |>
  fmt_percent(columns = c(share_2010, share_2024, change), decimals = 1) |>
  tab_source_note("Source: IPEDS undergraduate enrollment data aggregated to the state level.")
```
This state-level comparison highlights that gender composition has been shifting over time in many parts of the country. That broader context matters when interpreting institutional changes that might otherwise be attributed too quickly to a single national event.

# Op-Ed Calculation Section

```{r}
#| label: oped-calculations
# Baruch undergraduate body
baruch_ug <- IPEDS |>
  filter(str_detect(institution_name, "Baruch"), level == "all undergraduate") |>
  arrange(year)

baruch_fy <- IPEDS |>
  filter(str_detect(institution_name, "Baruch"), level == "first year undergrad") |>
  arrange(year)

baruch_ug_2024 <- baruch_ug |>
  filter(year == 2024) |>
  slice(1)

baruch_ug_total_2024 <- baruch_ug_2024$total_enrollment[[1]]
baruch_ug_white_2024 <- baruch_ug_2024$share_white[[1]]
baruch_ug_female_2024 <- baruch_ug_2024$share_female[[1]]
baruch_ug_div_2024 <- baruch_ug_2024$share_nonwhite_nonasian[[1]]

baruch_fy_2023 <- baruch_fy |>
  filter(year == 2023) |>
  slice(1)

baruch_fy_2024 <- baruch_fy |>
  filter(year == 2024) |>
  slice(1)

baruch_fy_total_2023 <- baruch_fy_2023$total_enrollment[[1]]
baruch_fy_total_2024 <- baruch_fy_2024$total_enrollment[[1]]
baruch_fy_div_2023 <- baruch_fy_2023$share_nonwhite_nonasian[[1]]
baruch_fy_div_2024 <- baruch_fy_2024$share_nonwhite_nonasian[[1]]
baruch_fy_white_2023 <- baruch_fy_2023$share_white[[1]]
baruch_fy_white_2024 <- baruch_fy_2024$share_white[[1]]

baruch_ug_2010 <- baruch_ug |>
  filter(year == 2010) |>
  slice(1)

baruch_ug_div_2010 <- baruch_ug_2010$share_nonwhite_nonasian[[1]]
baruch_ug_white_2010 <- baruch_ug_2010$share_white[[1]]

# California comparison institutions
berkeley_fy <- IPEDS |>
  filter(str_detect(institution_name, "University of California-Berkeley"),
         level == "first year undergrad",
         year %in% c(2023, 2024)) |>
  arrange(year)

sdsu_fy <- IPEDS |>
  filter(str_detect(institution_name, "San Diego State University"),
         level == "first year undergrad",
         year %in% c(2023, 2024)) |>
  arrange(year)

berkeley_2023_div <- berkeley_fy |>
  filter(year == 2023) |>
  pull(share_nonwhite_nonasian)

berkeley_2024_div <- berkeley_fy |>
  filter(year == 2024) |>
  pull(share_nonwhite_nonasian)

sdsu_2023_div <- sdsu_fy |>
  filter(year == 2023) |>
  pull(share_nonwhite_nonasian)

sdsu_2024_div <- sdsu_fy |>
  filter(year == 2024) |>
  pull(share_nonwhite_nonasian)
```

# Task 7: Op-Ed

## An Op-Ed from the President of Baruch College

As President of `r baruch_name`, I believe discussions about diversity should start with clear data. For this analysis, I define diversity as the share of undergraduate students who are neither White nor Asian American using IPEDS racial and ethnic categories. That definition is not perfect, but it provides a transparent way to track changes over time and compare institutions using consistent data.

In 2024, Baruch enrolled `r comma(baruch_ug_total_2024)` undergraduate students. Of those students, `r percent(baruch_ug_white_2024, accuracy = 0.1)` were classified as White, `r percent(baruch_ug_female_2024, accuracy = 0.1)` were women, and `r percent(baruch_ug_div_2024, accuracy = 0.1)` fell into my working definition of diversity. These numbers reflect the role Baruch plays as a large public institution serving students from across New York City.

Many observers have asked whether the SFFA decision produced an immediate shift in enrollment. At Baruch, our first-year class changed from `r comma(baruch_fy_total_2023)` students in 2023 to `r comma(baruch_fy_total_2024)` in 2024. The share of first-year students in my diversity measure moved from `r percent(baruch_fy_div_2023, accuracy = 0.1)` to `r percent(baruch_fy_div_2024, accuracy = 0.1)`, while the White share moved from `r percent(baruch_fy_white_2023, accuracy = 0.1)` to `r percent(baruch_fy_white_2024, accuracy = 0.1)`. These relatively small changes suggest that any immediate effect at Baruch was modest.

Long-term trends also matter. In 2010, the share of Baruch undergraduates captured by this diversity measure was `r percent(baruch_ug_div_2010, accuracy = 0.1)`. By 2024 it had risen to `r percent(baruch_ug_div_2024, accuracy = 0.1)`, while the White share fell from `r percent(baruch_ug_white_2010, accuracy = 0.1)` to `r percent(baruch_ug_white_2024, accuracy = 0.1)`. When compared with institutions such as UC Berkeley and San Diego State, which have operated without affirmative action for decades, these results suggest that year-to-year changes are often gradual and influenced by many factors beyond a single policy decision.

# Conclusion

This mini-project shows how a fairly simple descriptive workflow can still generate useful institutional insight. By combining enrollment data, institutional identifiers, and clear summary measures, I was able to answer targeted questions about Baruch, CUNY, California public universities, and national state-level patterns.

The main substantive takeaway is that 2024 enrollment shifts should be interpreted cautiously. Some institutions may show changes after SFFA, but meaningful interpretation requires comparison to longer-term trends and to benchmark institutions that were already operating under different admissions constraints. That broader context is what turns raw numbers into analysis.

# AI Usage Statement

I used Generative AI in a limited way to help me understand the assignment structure, plan the coding workflow, and troubleshoot R syntax. All written narrative text, analysis interpretation, and final op-ed language were written and edited by me.